---
title: 'Analysis of impact of weather events across US.'
author: "Pedro A. Alonso Baigorri"
date: "16 de junio de 2017"
output: html_document
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='figure/',
                      echo=TRUE, warning=FALSE, message=FALSE)
```

## Synopsis

Storms and other severe weather events can cause both public health and economic 
problems for communities and municipalities. Many severe events can result in 
fatalities, injuries, and property damage, and preventing such outcomes to the 
extent possible is a key concern.

This analysis involves exploring the U.S. National Oceanic and Atmospheric 
Administration's (NOAA) storm database. This database tracks characteristics of 
major storms and weather events in the United States, including when and where 
they occur, as well as estimates of any fatalities, injuries, and property damage.

The purpose of this analysis is to find out what are the type of weather events with 
more impact across the United States. The analysis of the impact will include, 
impact in the general population (including injuries and fatalities), and also 
economic impact.

This document will contain not only the analysis done but also the code and data
required to this analyis. The reason for this is to make it reproducible for 
other researchers.

Final conclusions will be supported by plots as a result of the analysis performed.

All the analysis has been done using R.

## Data processing

### Initial setup
First step for data processing is to load the main packages and setting my 
working environment.

```{r}
# setting the working path
setwd(".")


# Load required libraries
library(dplyr)
library(stringr)
library(ggplot2)
```

###Getting the data

Now, the next steps is to obtain and load  the data from the official repositority at:
https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

```{r cache=TRUE}
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
localFile <- "./data/storm.zip"
destFile <- "./data/storm.csv"


if (!file.exists(localFile))
{
    if (!file.exists("./data")){dir.create("./data")}
    
    download.file(fileURL, localFile)
    bunzip2(localFile, destFile, remove = FALSE )
}

storms <- read.csv(destFile)

dim(storms)
str(storms)

```

### Subsetting the dataset

According to the documentation the official event Types were changed in Jan 1996.
Before of this change, the number of different types was very small. So, I'll focus the
analysis only in the events occured after Jan 1996.

In addition I will select only the columns of interest for the analysis:

EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP

```{r}

# filtering data before of jan 1996
head(storms$BGN_DATE)
storms$date <- as.Date(storms$BGN_DATE, "%m/%d/%Y")
head(storms$date)
storms <- subset(storms, date > as.Date("12/31/1995", "%m/%d/%Y"))
dim(storms)
min(storms$date)

#selecting only the columns of interest
myvars <- c("EVTYPE", "FATALITIES", "INJURIES", 
             "PROPDMG", "PROPDMGEXP", "CROPDMG",
             "CROPDMGEXP")
storms <- subset(storms, select = myvars)
colnames(storms) <- myvars

dim(storms)

```

### Fixing the typos of event type column

I have observed that the values in the column EVTYPE have different typo erros such
as uppercases and whitespaces. So some basic operations have performed to fix these
problems.

```{r}
length(unique(storms$EVTYPE))

# transforming to uppercase
storms$EVTYPE <- toupper(storms$EVTYPE)
length(unique(storms$EVTYPE))

#removing starting white spaces
storms$EVTYPE <- trimws(storms$EVTYPE)
length(unique(storms$EVTYPE))
```


### Setting the right value to damage variables

The values of the economic damages are affected by an multiplier included in the 
columns PROPDMGEXP and CROPDMGEXP (for damages in PROPERTIES and DROPS).

The value of the multiplier is obtained from the following article:

https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html

The next code shows the processing required to calculate the total value with the 
corresponding multiplier.


```{r}

levels(storms$PROPDMGEXP)

#symbols are multipler of 0 except + that is 1
storms$PROPDMGEXP_VALUE[storms$PROPDMGEXP == ""] <- 0
storms$PROPDMGEXP_VALUE[storms$PROPDMGEXP == "-"] <- 0
storms$PROPDMGEXP_VALUE[storms$PROPDMGEXP == "?"] <- 0
storms$PROPDMGEXP_VALUE[storms$PROPDMGEXP == "+"] <- 1

# numbers from 0-9 are multiplier of 10 according to coursera forum article
# https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html
storms$PROPDMGEXP_VALUE[storms$PROPDMGEXP == "0"] <- 10
storms$PROPDMGEXP_VALUE[storms$PROPDMGEXP == "1"] <- 10
storms$PROPDMGEXP_VALUE[storms$PROPDMGEXP == "2"] <- 10
storms$PROPDMGEXP_VALUE[storms$PROPDMGEXP == "3"] <- 10
storms$PROPDMGEXP_VALUE[storms$PROPDMGEXP == "4"] <- 10
storms$PROPDMGEXP_VALUE[storms$PROPDMGEXP == "5"] <- 10
storms$PROPDMGEXP_VALUE[storms$PROPDMGEXP == "6"] <- 10
storms$PROPDMGEXP_VALUE[storms$PROPDMGEXP == "7"] <- 10
storms$PROPDMGEXP_VALUE[storms$PROPDMGEXP == "8"] <- 10
storms$PROPDMGEXP_VALUE[storms$PROPDMGEXP == "9"] <- 10

#then standard currency symbols
storms$PROPDMGEXP_VALUE[storms$PROPDMGEXP == "B"] <- 1000000000
storms$PROPDMGEXP_VALUE[storms$PROPDMGEXP == "h"] <- 100
storms$PROPDMGEXP_VALUE[storms$PROPDMGEXP == "H"] <- 100
storms$PROPDMGEXP_VALUE[storms$PROPDMGEXP == "K"] <- 1000
storms$PROPDMGEXP_VALUE[storms$PROPDMGEXP == "m"] <- 1000
storms$PROPDMGEXP_VALUE[storms$PROPDMGEXP == "M"] <- 1000

storms$PROPDMGEXP_VALUE <- storms$PROPDMGEXP_VALUE * storms$PROPDMG

#and now for CROPS
levels(storms$CROPDMGEXP)

storms$CROPDMGEXP_VALUE[storms$CROPDMGEXP == ""] <- 0
storms$CROPDMGEXP_VALUE[storms$CROPDMGEXP == "?"] <- 0
storms$CROPDMGEXP_VALUE[storms$CROPDMGEXP == "0"] <- 10
storms$CROPDMGEXP_VALUE[storms$CROPDMGEXP == "2"] <- 10
storms$CROPDMGEXP_VALUE[storms$CROPDMGEXP == "B"] <- 1000000000
storms$CROPDMGEXP_VALUE[storms$CROPDMGEXP == "k"] <- 1000
storms$CROPDMGEXP_VALUE[storms$CROPDMGEXP == "K"] <- 1000
storms$CROPDMGEXP_VALUE[storms$CROPDMGEXP == "m"] <- 1000
storms$CROPDMGEXP_VALUE[storms$CROPDMGEXP == "M"] <- 1000

storms$CROPDMGEXP_VALUE <- storms$CROPDMGEXP_VALUE * storms$CROPDMG

head(storms)

```



## Analysis

Once the preprocessing is finished the dataset is ready for the analysis.

The following analysis are performed:

- Analysis of fatalities
- Analysis of injuries
- Analysis of economic impact

### Analysis of fatalities

The following plot represent the top 10 types of events with more fatalites:

```{r}

fatalities <- aggregate(FATALITIES ~ EVTYPE, data = storms, FUN = sum )

fatalities <- fatalities[order(-fatalities$FATALITIES),]

top10 <- fatalities[1:10, ]


ggplot(top10, aes(x=reorder(EVTYPE, -FATALITIES), y=FATALITIES)) + geom_bar(stat="identity") + 
    labs(title="Top 10 Type of events with highest number of fatalites ", x="Type of Event", y="Number of fatalities")

fatalities_type <- top10[1, "EVTYPE"]
```


### Analysis of injuries

The following plot represent the top 10 types of events with more injuries:

```{r}

injuries <- aggregate(INJURIES ~ EVTYPE, data = storms, FUN = sum )

injuries <- injuries[order(-injuries$INJURIES),]

top10 <- injuries[1:10, ]

ggplot(top10, aes(x=reorder(EVTYPE, -INJURIES), y=INJURIES)) + geom_bar(stat="identity") + 
    labs(title="Top 10 Type of events with highest number of injuries ", x="Type of Event", y="Number of injuries")

injuries_type <- top10[1, "EVTYPE"]
```

### Analysis of economic damages

The following plot represent the top 10 types of events with more economic damages:

```{r}

storms$TOTAL_DAMAGES = (storms$PROPDMGEXP_VALUE + storms$CROPDMGEXP_VALUE)


eco <- aggregate(TOTAL_DAMAGES ~ EVTYPE, data = storms, FUN = sum )

eco <- eco[order(-eco$TOTAL_DAMAGES),]

top10 <- eco[1:10, ]

ggplot(top10, aes(x=reorder(EVTYPE, -TOTAL_DAMAGES), y=TOTAL_DAMAGES/1000000)) + 
    geom_bar(stat="identity") + 
    labs(title="Top 10 Type of events with highest number economic damages ", x="Type of Event", y="Damages (Millions of Dollars)")

economic_type <- top10[1, "EVTYPE"]
```

## Conclusions

The analysis concludes that:

- The event type with more impact in Injuries is: `r injuries_type`
- The event type with more impact in Fatalities is: `r fatalities_type`
- The event type with more Economic impact is: `r economic_type`

